---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r, echo = TRUE}
library(data.table)
data <- fread(unzip("./data/activity.zip"), sep=",", header=TRUE)
```

## What is mean total number of steps taken per day?
```{r, echo = TRUE, results="asis"}
library(xtable)
dailySteps <- as.vector(data[,sum(steps), by=date])
setnames(dailySteps, "V1", "steps")
hist(dailySteps[,steps], breaks=30, main="Total Daily Steps", xlab = "Steps", col="lightblue")
mean <- as.vector(data[,mean(steps), by=date])
setnames(mean, "V1", "mean")
median <- as.vector(data[,as.double(median(steps)), by=date])
setnames(median, "V1", "median")
xt <- xtable(merge(mean,median, by="date"))
print(xt, type="html")
```



## What is the average daily activity pattern?
```{r, echo=TRUE}
# sum steps by interval
intervalSteps <- data[,mean(steps, na.rm=TRUE), by=interval]
setnames(intervalSteps, "V1", "steps")
# Pad interval values and convert to POSIXct
intervalSteps$time <- as.POSIXct(strptime(sprintf("%04d",intervalSteps$interval), format="%H%M"), format = "%H:%M")
plot(intervalSteps$time, intervalSteps$steps, type="l", xlab="Time", ylab="Steps")
title("Mean Steps by Hour")
maxSteps <- cat('Max Steps at: ', strftime(unique(intervalSteps[steps==max(steps), time]), format="%H:%M"))
print(maxSteps, type='html')
```



## Imputing missing values
```{r, echo=TRUE, results="asis"}
# Total number of rows with missing values
incompleteRows <- cat('Rows with missing values: ', sum(is.na(data$steps)))
print(incompleteRows, type='html')

filledData <- merge(data, intervalSteps, by = "interval")
filledData[is.na(steps.x), steps.x := as.integer(steps.y)]
filledData[,steps.y:=NULL]
setnames(filledData, "steps.x", "steps")
filledData[order(date,interval)]

dailyFilledSteps <- as.vector(filledData[,sum(steps), by=date])
setnames(dailyFilledSteps, "V1", "steps")
hist(dailyFilledSteps[,steps], breaks=30, main="Total Daily Steps", xlab = "Steps", col="lightblue")
meanFilled <- as.vector(filledData[,mean(steps), by=date])
setnames(meanFilled, "V1", "mean")
medianFilled <- as.vector(filledData[,as.double(median(steps)), by=date])
setnames(medianFilled, "V1", "median")
xt <- xtable(merge(meanFilled, medianFilled, by="date"))
print(xt, type="html")
```



## Are there differences in activity patterns between weekdays and weekends?
```{r, echo=TRUE}
library(ggplot2)
daysOfWeek <- filledData[, weekdays(as.POSIXct(date))]
isWeekend <- grepl("\\<Saturday\\>|\\<Sunday\\>", daysOfWeek)
filledData$weekday <- factor(isWeekend, labels = c("weekday", "weekend"))
weekdayIntervalSteps <- filledData[weekday=='weekday',mean(steps, na.rm=TRUE), by=interval]
setnames(weekdayIntervalSteps, 'V1', 'steps')
weekdayIntervalSteps$weekday <- "weekday"

weekendIntervalSteps <- filledData[weekday=='weekend',mean(steps, na.rm=TRUE), by=interval]
setnames(weekendIntervalSteps, 'V1', 'steps')
weekendIntervalSteps$weekday <- "weekend"

weekendAverages <- rbind(weekendIntervalSteps, weekdayIntervalSteps)

ggplot(weekendAverages, aes(interval, steps)) +
        geom_line(color="steelblue") +
        facet_grid(weekday ~ .) +
        labs(title="Weekday and Weekend Average Steps by Time") +
        labs(y="Steps", x="Time")

